Relationship between solar wind, $D_{st}$, and plasmasphere trough mass density on one-hour time scales

### 2016-03-18 paper.tex
Introducton changes

There are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}), but and dependence of the plasma trough on geomagnetic and solar wind conditions has not been extensively studied until the past decade.

The dependence of the plasmapause location on geomagnetic activity was observed as early as \note{Cite}.  \cite{Gallagher1988EmpiricalModelPlasmasphere} used measurements from the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model plasmasphere density using a relationship dependent on $L$ shell, altitude, geomagnetic latitude, and observations grouped by a weighted average of $K_p$.  The developed statistical model showed significant improvements when a weighted average of $K_p$ over three days was used to group the observations used in the model.  \note{Discuss plasma trough relevance}.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (obervations were made when CRRESS was in an elliptical, near equatorial orbit). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$, corresponding to the plasma trough region.

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation between 27 day averages fo F10.7 and 27-day medians of $\rho_{eq}$.

@@

\cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-sheels (2-14) had a strong correlation (0.88).  Although the correlation in the plasma trough was not caluclated, the data presented show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The results in the papers discussed indicate that (1) the mass density in the plasma trough region is best correlated with $F10.7$ and (2) there exists a correlation between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify a statistical characterization of the relationship between geomagnetic activity and mass density.  In addition, we attempt to identify solar wind processes may drive this changes.

There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, as do geomagnetic and solar wind parameters (3) the dependence on the north/south component of the solar wind magnetic field is expected to be nonlinear. (For northward IMF and low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in density associated with strong southward IMF and enhanced magnetospheric activity [\cite{Takahashi2010}; \cite{Yao2008}]).

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of evens, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate F10.7 depenences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF and also by separation by averages of the northrward and southward IMF.

@@

>>>>

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966}.  There are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}, \cite{LemaireEarthsPlasmasphere}, \cite{Takahashi2006}, and \cite{Takahashi2010}), but and dependence of the plasma trough on geomagnetic and solar wind conditions has not been extensively studied until the past decade.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (obervations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having L between 5 and 7 $R_E$.

@@

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, \vinote{Connect to harmonic toroidal frequency} a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation ($\sim 0.93$) between 27-day averages of F10.7 and 27-day medians of $\rho_{eq}$.

@@

\cite{Yao2008} studied the relationship between $D_{st}$ and ion number density for low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-sheels (2-14) had a strong correlation (0.88).  Although the correlation in the plasma trough was not calculated, the data presented show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The results in the papers discussed indicate that (1) the mass density in the plasma trough region is best correlated with $F10.7$ and (2) there exists a correlation between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density in the plasma trough region.  In addition, we attempt to identify solar wind processes may drive changes in mass density.

There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, and geomagnetic and solar wind parameters also correlate with F10.7, (3) the dependence of mass density on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, corresponding tolow solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of mass density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in mass density associated with strong southward IMF and enhanced magnetospheric activity [\cite{Takahashi2010}; \cite{Yao2008}].

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 depenences.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northrward and southward IMF.

@@

\cite{Gallagher1988EmpiricalModelPlasmasphere} used measurements from the Retarding Ion Mass Spectrometer onboard the Dynamics Explorer 1 satellite to model plasmasphere density using a relationship dependent on $L$ shell, altitude, geomagnetic latitude, and observations grouped by a weighted average of $K_p$.  The developed statistical model showed significant improvements when a weighted average of $K_p$ over three days was used to group the observations used in the model.  \note{Discuss plasma trough relevance}.
